NMRlipids Databank: Making data-driven analyses of membrane properties accessible for all

### 2022-12-06 main.tex
Update main.tex

While the NMRlipids databank currently contains only bilayer systems, it can readily be expanded to contain other molecules than lipids, such as disordered proteins or sugars. 
%a solution for lipid bilayers based on overlay databank structure illustrated in Fig.~\ref{fig:overlay}.  
The combination of an overlay databank structure and a community-wide open collaboration used here has been applied successfully when building databanks in other fields,  
%is developed here to solve the practical challenges in generating databanks of MD simulation data enabling flexible analyses over large sets of simulation data, but it can be  potentially used for wide range of situations. 
particularly when storage of raw data requires significant resources, best practices in the field are not defined, and incentives to share data do not exist.
%overlay databank approach lowers the barrier to start without compromizing the long term stability or scalability.
Thus, the approach has proven itself in enabling data-driven and machine learning applications in new fields.

@@

%Up to three levels of \textbf{subheading} are permitted. Subheadings should not be numbered.
\subsection{Programmatic access to MD simulation data of membranes composed of most common biologically abundant lipids}
NMRlipids databank is a community driven databank containing atomistic MD simulations of biologically relevant lipid membranes emerging from the NMRlipids open collaboration~\cite{botan15,ollila16,catte16,antila19,bacle21}. Nearly thousand simulation trajectories with the total length approaching half a millisecond can be accessed programmatically using Python API or graphical user interface (\url{http://www.databank.nmrlipids.fi/}). 
%NMRlipids databank is the only databank of MD simulations with open programmatic access to its content enabling fully flexible analyses of its content. 
Following the FAIR principles~\cite{wilkinson16} and NMRlipids project protocol, not only the content and computer programs related to the databank but also the whole construction process of the databank are open access~\cite{botan15}. 
%The NMRlipids databank will be always open for data and other contributions.

Currently available single component lipid membranes and binary mixtures in the NMRlipids databank are illustrated in Fig.~\ref{fig:overlay}D. While these compose the majority of available trajectories, also mixtures up to five lipids are available. The distribution of available simulations containing a specific lipid is shown in Fig.~\ref{fig:overlay}C. This roughly resembles the biological abundancies of different lipid types, PC being the most common  followed by cholesterol, PE, PS, PG, PI, and other lipids depending on organism and organelle~\cite{vanmeer08}. The force fields used in the simulations cover all the essential parameters used in lipid simulations, ranging from highly popular CHARMM36 parameters to united atom and polarizable force fields. Therefore, the averages calculated over the databank can be considered as mean predictions from available lipid models for an average cell membrane.
%Currently the databank is composed of approximately 500 trajectories with the total length of approximately 231 microseconds of which most are contributed for the previous publications from the NMRlipids open collaboration~\cite{botan15,catte16,antila19,bacle21}. The distribution of lipids, force fields, length of the trjajectories and available binary mixtures are shown in Fig.~\ref{fig:overlay}.

The efficient upcycling of large MD simulation trajectories in the NMRlipids databank is enabled by the overlay structure illustrated in Fig.~\ref{fig:overlay}A. Raw simulation data in the {\it data layer} can be stored in any publicly available location with long term stability and permanent links to the data, such as digital object identifiers. The {\it Databank layer} is the core of the databank containing all the relevant information about the simulations, including links to the raw data, relevant metadata describing the systems, universal naming conventions for lipids and their atoms, computer programs to create and analyse the entries, basic properties calculated from all simulations (area per lipid, C-H bond order parameters, x-ray scattering form factors and membrane thickness), and quality evaluation of simulations against experimental data. In practise, this information is stored is the git repository which is currently available at~\url{https://github.com/NMRLipids/Databank}. The {\it application layer} contains outputs from the databank, such as graphical user interface (\url{http://www.databank.nmrlipids.fi/}) and results from analyses described in sections below. More detailed description on the NMRlipids databank structure is available the supplementary information.

@@

\subsection{Selection of best simulation parameters for applications using the NMRlipids databank}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
Understanding the accuracy of used force field parameters is crucial when estimating the reliability and significance MD simulation results. However, the lack of universal quality measures for MD simulations and complicated landscape of force field quality for membranes hamper the quality of simulations which has resulted to many controversial results~\cite{antila22b}. For example, previous quality evaluations of membrane simulations have concluded that CHARMM36 parameters give the best description for lipid headgroup conformational ensembles~\cite{bacle21}, GROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}, and that OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. 
%, thereby being a major obstacle in many applications of membrane MD simulations. 
To enable the navigation in this complex force field quality landscape, we define here quantitative quality measures that can be used to rapidly find the best available parameters for a specific application or to guide force field development. For detailed definitions of quality measures, see the supplementary information.

@@

Ranking of all simulations with the highest scores based on evaluation against C-H bond order parameters from NMR experiments is shown in Fig.~\ref{fig:quality} A and ranking of best POPE lipid bilayer simulations in Fig.~\ref{fig:quality} B. Direct comparisons with experiments for selected simulations are shown in Figs.~\ref{fig:quality} C-D. C-H bond order parameters are sensitive to conformational ensembles of individual lipid molecules~\cite{ollila16}, but they correlate well also with the membrane thickness and lateral packing as shown in Fig.~\ref{fig:quality} F. This is complemented by comparing the absolute values of x-ray scattering form factors between simulations and experiments that are related to the overall membrane dimensions via electron density profile. Particularly, the location of minima in these form factors are correlated with the membrane dimensions as shown in Figs.~\ref{fig:quality} F,~\ref{fig:QualityCorrelationsSI}, and~\ref{fig:sizedependence}.

>>>>

While the NMRlipids databank currently contains only bilayer systems, it can readily be expanded to contain other molecules than lipids, such as disordered proteins or sugars. 
%a solution for lipid bilayers based on overlay databank structure illustrated in Fig.~\ref{fig:overlay}.  
The combination of an overlay databank structure and a community-wide open collaboration used here has been applied successfully when building databanks in other fields,  
%is developed here to solve the practical challenges in generating databanks of MD simulation data enabling flexible analyses over large sets of simulation data, but it can be  potentially used for wide range of situations. 
particularly when storage of raw data requires significant resources, best practices in the field are not defined, and incentives to share data do not exist.
%overlay databank approach lowers the barrier to start without compromizing the long term stability or scalability.
Thus, the approach has already proven itself most useful in enabling data-driven and machine learning applications in new fields.

@@

%Up to three levels of \textbf{subheading} are permitted. Subheadings should not be numbered.
\subsection{Programmatic access to MD simulation data of membranes composed of most common biologically abundant lipids}
NMRlipids databank is a community-driven databank containing atomistic MD simulations of biologically relevant lipid membranes emerging from the NMRlipids open collaboration~\cite{botan15,ollila16,catte16,antila19,bacle21}. Nearly a thousand simulation trajectories with the total length approaching half a millisecond can be accessed programmatically using the Python application programming interface (API) or a graphical user interface (GUI) (\url{http://www.databank.nmrlipids.fi/}). 
%NMRlipids databank is the only databank of MD simulations with open programmatic access to its content enabling fully flexible analyses of its content. 
Following the FAIR principles~\cite{wilkinson16} and NMRlipids project protocol, not only the content and computer programs related to the databank, but also the entire construction process of the databank, are open access~\cite{botan15}. 
%The NMRlipids databank will be always open for data and other contributions.

Currently available single component lipid membranes and binary mixtures in the NMRlipids databank are illustrated in Fig.~\ref{fig:overlay}D. While these comprise the majority of available trajectories, mixtures with up to five lipid types are available. The distribution of available simulations containing a specific lipid is shown in Fig.~\ref{fig:overlay}C. This roughly resembles the biological abundance of different lipid types with phosphatidylcholine (PC) being the most common followed by cholesterol, phosphatidylethanolamine (PE), phosphatidylserine (PS), phosphatidylglycerol (PG), phosphatidylinositol (PI), and other lipids depending on organism and organelle~\cite{vanmeer08}. The force fields used in the simulations cover all the essential parameters commonly used in lipid simulations, ranging from highly popular CHARMM36 parameters to united atom and polarizable force fields. Therefore, the averages calculated over the databank can be considered as mean predictions from available lipid models for an average cell membrane.
%Currently the databank is composed of approximately 500 trajectories with the total length of approximately 231 microseconds of which most are contributed for the previous publications from the NMRlipids open collaboration~\cite{botan15,catte16,antila19,bacle21}. The distribution of lipids, force fields, length of the trjajectories and available binary mixtures are shown in Fig.~\ref{fig:overlay}.

The efficient upcycling of large MD simulation trajectories in the NMRlipids databank is enabled by the overlay structure illustrated in Fig.~\ref{fig:overlay}A. Raw simulation data in the {\it data layer} can be stored in any publicly available location with long term stability and permanent links to the data (e.g., digital object identifiers (DOIs)). The {\it Databank layer} is the core of the databank containing all the relevant information about the simulations, including links to the raw data, relevant metadata describing the systems, universal naming conventions for the lipids and their atoms, computer programs to create and analyse the entries, basic properties calculated from all simulations (area per lipid, C-H bond order parameters, X-ray scattering form factors and membrane thickness), and quality evaluation of simulations against experimental data. In practise, this information is stored is the git repository which is currently available at~\url{https://github.com/NMRLipids/Databank}. The {\it application layer} contains outputs from the databank, such as graphical user interface (\url{http://www.databank.nmrlipids.fi/}), and results from analyses described in sections below. A more detailed description on the NMRlipids databank structure is available the supplementary information.

@@

\subsection{Selection of best simulation parameters for applications using the NMRlipids databank}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
Understanding the accuracy of used force field parameters is crucial when assessing the reliability and significance MD simulation results. However, the lack of universal quality measures for MD simulations and the complicated landscape of force field quality for membranes hamper the applicability such simulations, resulting in substantial controversies~\cite{antila22b}. For example, previous quality evaluations of membrane simulations have concluded that CHARMM36 parameters give the best description for lipid headgroup conformational ensembles~\cite{bacle21}, GROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}, and that OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. 
%, thereby being a major obstacle in many applications of membrane MD simulations. 
To enable the navigation in this complex force field milieu, we here define quantitative quality measures that can be used to rapidly find the best available parameters for a specific application or to guide the development of improved force fields. For detailed definitions of quality measures, see the supplementary information.

@@

Ranking of all simulations with the highest scores based on evaluation against C-H bond order parameters from NMR experiments is shown in Fig.~\ref{fig:quality} A and ranking of best POPE lipid bilayer simulations in Fig.~\ref{fig:quality} B. Direct comparisons with experiments for selected simulations are shown in Figs.~\ref{fig:quality} C-D. C-H bond order parameters are sensitive to conformational ensembles of individual lipid molecules~\cite{ollila16} while also correlating well with the membrane thickness and lateral packing as shown in Fig.~\ref{fig:quality} F. This is complemented by comparing the absolute values of X-ray scattering form factors between simulations and experiments that are related to the overall membrane dimensions via the electron density profile. In particular, the location of minima in these form factors are correlated with the membrane dimensions as shown in Figs.~\ref{fig:quality} F,~\ref{fig:QualityCorrelationsSI}, and~\ref{fig:sizedependence}.
